---
title: "regression_final"
author: "Zihan Wu"
date: "2023-12-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
library(stats)
library(boot)
library(caret)
set.seed(2023)
```

### Data Loading and Preprocessing
```{r}
# Load the rat sightings data
rat_data <- read_csv("data/rat_sightings.csv")

# Convert 'Created Date' to datetime format and extract year and month
rat_data$Created_Date <- as.Date(rat_data$`Created Date`, format="%m/%d/%Y %H:%M:%S")
rat_data$Year <- year(rat_data$Created_Date)
rat_data$Month <- month(rat_data$Created_Date)

# Define start dates for Rat Czar and COVID-19 pandemic
rat_czar_start_date <- as.Date("2023-04-12")
pandemic_start_date <- as.Date("2020-03-11")

# Creating binary variables for pre and post Rat Czar and COVID-19 periods
rat_data$Pre_Rat_Czar <- ifelse(rat_data$Created_Date < rat_czar_start_date, 1, 0)
rat_data$Post_Rat_Czar <- ifelse(rat_data$Created_Date >= rat_czar_start_date, 1, 0)
rat_data$Pre_Covid <- ifelse(rat_data$Created_Date < pandemic_start_date, 1, 0)
rat_data$During_Covid <- ifelse(rat_data$Created_Date >= pandemic_start_date & rat_data$Created_Date < rat_czar_start_date, 1, 0)
rat_data$Post_Covid <- ifelse(rat_data$Created_Date >= rat_czar_start_date, 1, 0)

# Aggregate data by month and year for rat sightings
monthly_sightings <- rat_data %>%
  group_by(Year, Month) %>%
  summarise(Sightings = n())

# Merge with original data to include binary variables
merged_data <- merge(monthly_sightings, unique(rat_data[,c("Year", "Month", "Pre_Rat_Czar", "Post_Rat_Czar", "Pre_Covid", "During_Covid", "Post_Covid")]), by = c("Year", "Month"))
```

### Regression Analysis

#### Impact of Rat Czar Appointment

```{r}
model_rat_czar <- lm(Sightings ~ Pre_Rat_Czar, data = merged_data)
summary(model_rat_czar)
```

#### Impact of COVID-19 Pandemic

```{r}
model_covid <- lm(Sightings ~ Pre_Covid + During_Covid, data = merged_data)
summary(model_covid)
```
### Interpretation of Regression Analysis Results

#### Impact of Rat Czar Appointment

The regression analysis indicates a significant increase in rat sightings following the appointment of the Rat Czar. The model, with an R-squared value of 0.112, explains approximately 11.2% of the variance in rat sightings. The significant negative coefficient for `Pre_Rat_Czar` (-936, p < 0.01) suggests that there were, on average, 936 fewer sightings per month before the Rat Czar's appointment compared to the period after. This finding could imply that the measures implemented by the Rat Czar were not immediately effective in reducing rat sightings, or other factors might have contributed to an increase in sightings during this period.

#### Impact of COVID-19 Pandemic

The COVID-19 pandemic also appears to have had a significant impact on rat sightings. The model's R-squared value of 0.3953 indicates that it explains about 39.53% of the variance in rat sightings. The coefficients for `Pre_Covid` and `During_Covid` are -1116.1 (p < 0.01) and -353.2 (p = 0.0531), respectively. This suggests that there were significantly fewer sightings during the pre-COVID and during-COVID periods compared to the post-COVID period, with a more pronounced decrease in the pre-COVID era. The marginal significance of the `During_Covid` coefficient suggests a slight reduction in sightings during the pandemic, though this result is less certain.

#### Overall Conclusion

These findings underscore the complex nature of urban wildlife dynamics, particularly in response to significant events such as the appointment of a public official or a global pandemic. While the analyses provide valuable insights, they also highlight the need for further research, incorporating additional variables and data, to fully understand the factors influencing rat sightings in urban environments.

### Bootstrapping of Rat Czar

```{r}
# Define the statistic function for bootstrapping with only Pre_Rat_Czar
stat_function <- function(data, indices) {
  d <- data[indices, ] # allows bootstrapping to select sample
  fit <- lm(Sightings ~ Pre_Rat_Czar, data = d)
  return(coef(fit))
}

# Apply the bootstrapping with R = 1000 bootstrap replicates
results <- boot(data = merged_data, statistic = stat_function, R = 1000)

# Print the results
print(results)
```
### Interpretation of Bootstrap Results

The bootstrapping process was applied to the regression analysis of rat sightings with respect to the Rat Czar appointment. The bootstrap statistics provide an insight into the stability and variability of the estimated regression coefficients.

#### Bootstrap Results:

- **Intercept (Post_Rat_Czar period) Coefficient**:
  - **Original**: 2285.875
  - **Bias**: -5.886
  - **Standard Error**: 173.254

- **Pre_Rat_Czar Coefficient**:
  - **Original**: -936.043
  - **Bias**: 3.779
  - **Standard Error**: 176.354

#### Conclusions from Bootstrap Analysis:

The bootstrap analysis shows that the bias for both coefficients is small, which suggests that the estimates from our original model are stable across different samples of our data. The standard errors from the bootstrap procedure provide a measure of the variability in the estimates and can be used to construct more robust confidence intervals that do not rely on the normality assumption of the residuals.

### Cross-Validation of Rat Czar

```{r}
# Define control using k-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Train the model
model <- train(Sightings ~ Pre_Rat_Czar + Post_Rat_Czar,
               data = merged_data,
               trControl = train_control,
               method = "lm")

# Summarize the results
print(model)
```
### Bootstrap Analysis for COVID-19 Impact

```{r}
# Define the statistic function for bootstrapping
stat_function_covid <- function(data, indices) {
  d <- data[indices, ] # allows bootstrapping to select sample
  fit <- lm(Sightings ~ Pre_Covid + During_Covid, data = d)
  return(coef(fit))
}

# Apply the bootstrapping with R = 1000 bootstrap replicates
bootstrap_results_covid <- boot(data = merged_data, statistic = stat_function_covid, R = 1000)

# Print the results
print(bootstrap_results_covid)
```
### Cross-Validation for COVID-19 Impact

```{r}
# Define the control using 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Train the model
model_cv_covid <- train(Sightings ~ Pre_Covid + During_Covid,
                        data = merged_data,
                        trControl = train_control,
                        method = "lm")

# Summarize the results
print(model_cv_covid)
```

### Interpretation of Bootstrap and Cross-Validation Results for COVID-19 Impact Analysis

We performed a bootstrap analysis and cross-validation to assess the stability and predictive performance of our regression model, which estimates the impact of the COVID-19 pandemic on rat sightings.

#### Bootstrap Results for COVID-19 Analysis

The bootstrap analysis provides an estimate of the variability of the regression coefficients:

- **Intercept (Post-COVID period) Coefficient**:
  - **Original**: 2285.875
  - **Bias**: -3.056
  - **Standard Error**: 172.188

- **Pre_COVID Coefficient**:
  - **Original**: -1116.111
  - **Bias**: 3.621
  - **Standard Error**: 175.594

- **During_COVID Coefficient**:
  - **Original**: -353.191
  - **Bias**: 1.357
  - **Standard Error**: 193.458

The small biases suggest that the coefficient estimates are robust and not highly sensitive to the specific sample of data used. The standard errors reflect the variability in the coefficients and can be used to construct confidence intervals. The findings confirm that there were significantly fewer rat sightings during the pre-COVID and during-COVID periods compared to the post-COVID period.

#### Cross-Validation Results for COVID-19 Analysis

Cross-validation was used to assess the model's predictive performance. The results are as follows:

- **RMSE (Root Mean Squared Error)**: 470.6385
- **R-squared**: 0.3918
- **MAE (Mean Absolute Error)**: 382.339

The RMSE and MAE provide measures of the model's prediction error, with lower values indicating better fit. The R-squared value indicates that approximately 39.18% of the variance in rat sightings is explained by the model across the cross-validated datasets.

These results suggest that while the model has a moderate explanatory power, there is still a considerable amount of variability in rat sightings that the model does not capture. This could be due to other factors not included in the model or inherent variability in the data.

Overall, the analysis suggests that the COVID-19 pandemic had a statistically significant impact on rat sightings, with changes in human behavior during the pandemic likely contributing to fluctuations in rat sighting reports.











